Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-10 paper/paper.tex
Updates from Overleaf

%%%%%% Comment related
\newcommand{\red}[1]{[{\bf  \color{red}{ST: #1}}]}
\newcommand{\xred}[1]{[{\bf  \color{red}{\sout{ST: #1}}}]}
\newcommand{\green}[1]{[{\bf  \color{green}{XG: #1}}]}
\newcommand{\xgreen}[1]{[{\bf  \color{green}{\sout{XG: #1}}}]}
\newcommand{\blue}[1]{[{\bf  \color{blue}{JG: #1}}]}
\newcommand{\xblue}[1]{[{\bf  \color{blue}{\sout{JG: #1}}}]}
\newcommand{\material}[1]{\iffalse[{\bf  \color{cyan}{Material: #1}}]\fi}

@@

\begin{abstract}
We introduce a method of using generic programming tensor network to compute various properties of independent sets,
which include, for example, the size of the maximum independent sets and the number and enumeration of independent sets of a given size.
By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the treewidth of the problem graph.
Our framework can be easily extended to compute properties of other problems such as the cut size, coloring, and maximal cliques, among others.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there is an important class of problems that can be quantified locally,
including the independent set problem, the cutting problem, the 3-coloring problem, the 3-satisfiability problem,
the maximal clique problem, and the vertex cover problem et al ~\cite{Moore2011}.
There is a widespread interest of these problem because they have wide range of applications in scheduling, logistics,
wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}.
\blue{need more diversed applications not limited to IS}

The most studied version of the above problems is finding the maximum/minimum set size and one of the solution with that size satisfying the contraints,
while many of them belong to the hardness category NP-hard~\cite{Hastad1996}, i.e. there is unlikely a polynomial time algorithm for solving them.
In this paper, we focus on the less studied \textit{property} computation of these problems which can be harder but is crutial towards understanding a problem.
The property computation includes finding the maximum/minimum set size, counting sets of a give size, enumerating all such sets or direct sampling them when they are not enumeratable.
For the weighted version, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.

We provide an abstract algebra view for the computation of all these properties and proposed a unified framework,
generic tensor-network, to solve them all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network or einsum in different fields,
where its tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
While generic tensor network is a generalization of which by extending the tensor element type to a generic one being a commutative semiring.
%Traditional algorithms for finding the MIS size include the branching algorithms~\cite{Tarjan1977, Robson1986} and dynamic programming based algorithms~\cite{Courcelle1990, Fomin2013}.
Given a good contraction order, contracting a tensor network in general has time complexity $2^{{\rm tw}(G)}$, where ${\rm tw}(G)$ is the treewidth of graph $G$.
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find the optimal contraction order of a large scale tensor network in reasonable time.
Although this time complexity is just on par with the dynamic programming algorithms~\cite{Courcelle1990, Fomin2013} for solving the maximum size and one of the best set, tensor network has richer algebraic structure, and this structure is crutial for computing properties.
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

To avoid the discussion being overly diversed, we limit the target problem to the independent set problem,
and discuss other combinatorial optimization problems in \App{app:otherproblems}.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$,
and we denote the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$. 
%The independent set problem is related to physics applications such as in the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
%in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022};
%Property computation can, for example, be used to understand phase transitions~\cite{Dyre2016, Fernandes2007},
%to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the presence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} related to the clustering phenomenon of large independent sets.
Meanwhile, finding the number of independent sets of a given size is equivalent to computing the coefficients of the independence polynomial~\cite{Harvey2018,Ferrin2014}.
which is a useful graph characteristic related to, for example, the partition functions~\cite{Lee1952,Yang1952} and Euler characteristics of the independence complex~\cite{Bousquet2008, Levit2009}.
%However, there is a lack of general and versatile tool to compute these different properties of independent sets.

The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce an independent set problem to a tensor network contraction problem \Sec{sec:tnmap}
and show how to engineer the element types in \Sec{sec:counting},\Sec{sec:enumeration} and \Sec{sec:weighted} to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we benchmark our algorithms in \Sec{sec:benchmark} and provide a few examples in \Sec{sec:examples} to demonstrate the versatility of our tool.
We show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices
and analyzing the presence of overlap gap property on three regular graphs.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

\section{Tensor networks}\label{sec:tn}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts;
it can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in the subscript to represent a degree of freedom.
One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice.
Hence a tensor network can be represented as a simple graph,
where a tensor is mapped to a vertex and a label shared by two tensors is mapped to an edge.
In this work, we do not impose such a restriction, so an index can appear an arbitrary number of times. 
The graphical representation of a generalized tensor network is a hypergraph, in which an edge (a label in subscript) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{app:tensorbad}.

@@

\section{Generic programming tensor contractions}\label{sec:generic}
In previous works relating tensor networks and combinatoric problems~\cite{Kourtis2019, Biamonte2017},
the elements in the tensor networks are limited to standard number types such as floating point numbers and integers.
Owning to the development of modern compiling technology, we no longer need to limit our imagination to standard number types.
One of the key concepts that push the technology forward is called generic programming:

@@

\section{Tensor network representation of independent sets} \label{sec:tnmap}
Let $G=(V,E)$ be a weighted graph with each vertex $i\in V$ being associated with a weight $w_i$.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i, w_i)$ as
\begin{equation}
    W(x_i, w_i) = \left(\begin{matrix}
        1 \\
        x_i^{w_i}
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i, w_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i, w_i)_1=x_i^{w_i}$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i, w_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.

@@

    \centerline{\begin{tikzpicture}[
    dot/.style = {circle, fill, minimum size=#1,
                inner sep=0pt, outer sep=0pt},
    dot/.default = 6pt  % size of the circle diameter 
                    ]  
        \def\dx{0};
        \def\r{0.25cm}
        \filldraw[fill=black] (\dx,0) circle [radius=\r];
        \filldraw[fill=black] (\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,0) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (2.5+\dx,2.5) circle [radius=\r];
        \draw [black,thick] (\dx,0) -- (\dx,1.5);
        \draw [black,thick] (\dx,0) -- (1.5+\dx,0);
        \draw [black,thick] (\dx,1.5) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (\dx,1.5);
        \draw [black,thick] (2.5+\dx,2.5) -- (1.5+\dx,1.5);
        \node[color=white] at (\dx,0) {a};
        \node[color=white] at (\dx,1.5) {b};
        \node[color=white] at (1.5+\dx,0) {c};
        \node[color=white] at (1.5+\dx,1.5) {d};
        \node[color=white] at (2.5+\dx,2.5) {e};
        \def\dx{5};
        \def\r{0.25cm}
        \foreach \x/\y/\e in {0.75/0/ac, 0/0.75/ab, 1.5/0.75/cd, 0.75/1.5/bd, 0.75/0.75/bc, 2/2/de}
            \node[color=white,fill=black,dot=2*\r] at (\x+\dx,\y) (\e) {\scriptsize B};
        \foreach \x/\y/\v in {0/0/a, 0/1.5/b, 1.5/0/c, 1.5/1.5/d, 2.5/2.5/e}
            \node[color=black] at (\x+\dx,\y) (\v) {$s_\v$};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=white,fill=black,dot=\r] at (\x+\dx,\y) (\v\v) {};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=black] at (\x+\dx+0.6,\y) {\scriptsize $W(x_\v, w_\v)$};
        \draw [cyan,thick] (a) -- (aa);
        \draw [cyan,thick] (a) -- (ab);
        \draw [cyan,thick] (a) -- (ac);
        \draw [blue,thick] (b) -- (bb);
        \draw [blue,thick] (b) -- (ab);
        \draw [blue,thick] (b) -- (bc);
        \draw [blue,thick] (b) -- (bd);
        \draw [red,thick] (c) -- (cc);
        \draw [red,thick] (c) -- (ac);
        \draw [red,thick] (c) -- (bc);
        \draw [red,thick] (c) -- (cd);
        \draw [green,thick] (d) -- (dd);
        \draw [green,thick] (d) -- (bd);
        \draw [green,thick] (d) -- (de);
        \draw [green,thick] (d) -- (cd);
        \draw [orange,thick] (e) -- (ee);
        \draw [orange,thick] (e) -- (de);
    \end{tikzpicture}}

    The contraction of this tensor network can be done in a pairwise order utilizing the associativity, additive commutativity, and multiplicative commutativity of tensor elements:
    \small{
    \begin{align*}
        &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a, w_a)_{s_a} W(x_b, w_b)_{s_b} W(x_c, w_c)_{s_c} W(x_d, w_d)_{s_d} W(x_e, w_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_c s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}\\
        =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e,w_e)_{s_e}\right) W(x_d,w_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b, w_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c, w_c)_{s_c}\right)\right)\right.\\
        &\phantom{XXX}\left.\left(B_{s_bs_c}\left(\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a,w_a)_{s_a}\right)\right)\right)\right)\\
        =&1 + x_a^{w_a} + x_b^{w_b} + x_c^{w_c} + x_d^{w_d} + x_e^{w_e} + x_a^{w_a}x_d^{w_d} + x_a^{w_a}x_e^{w_e} + x_c^{w_c}x_e^{w_e} + x_b^{w_b}x_e^{w_e}\\
        =&1+5x+4x^2 \qquad \quad (x_{i} = x, w_{i}=1).
    \end{align*}}
\end{example}

\section{Independence polynomial}\label{sec:indpoly}
In the special case of $x_i = x$ and $w_i = 1$, the contraction result directly corresponds to the independence polynomial.
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
where $a_k$ is the number of independent sets of size $k$ in graph $G=(V,E)$. The total number of independent sets is thus equal to $I(G, 1)$.
Its connection to \Eq{eq:idp} can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise. The summation counts the number of independent sets of size $k$. 
One can feed the standard computer numbers types into this tensor network and evaluate the independence polynomial by contracting it numerically.
However, we are more interested in finding the coefficients of this polynomial, because this quantity tells us the counting of independent sets at each size.
To achieve this, let us first elevate the tensor elements $0$s and $1$s in tensors $W(x, 1)$ and $B$ from integers and floating point numbers to the additive identity,

@@

In practise, using the polynomial type suffers a space overhead proportional to $\alpha(G)$ because each polynomial requires a vector of such size to store the coefficients. 

@@

\section{Enumerating and sampling independent sets}\label{sec:enumeration}
\subsection{Set algebra for configuration enumeration}
The enumeration problems of independent sets are also interesting and has been studied extensively in the literature~\cite{Bron1973, Eppstein2010, Johnson1988}, including,
for example, the enumeration of all independent sets, the enumeration of all maximal independent sets, or the enumeration of all MISs.
To enumerate all independent sets, we designed an algebra defined on sets of bitstrings:
\begin{equation}
\eqname{SN}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau \, | \, \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}.
\end{split}\label{eq:set}
\end{equation}
where $s$ and $t$ are each a set of $|V|$-bit strings and $\lor^\circ$ is the bitwise OR operation over two bit strings.
\begin{example}\label{eg:setalgebra}
    For elements being bit strings of length $5$, we have the following set algebra
\begin{equation*}
\begin{split}
    &\{00001\} \oplus \{01110, 01000\} = \{01110, 01000\} \oplus \{00001\} = \{00001,01110, 01000\}\\
    &\{00001\} \oplus \{\} = \{00001\}\\
&\\
    &\{00001\} \odot \{01110, 01000\} = \{01110, 01000\} \odot \{00001\} = \{01111, 01001\}\\
    &\{00001\} \odot \{\} = \{\}\\
    &\{00001\} \odot \{00000\} = \{00001\}.
\end{split}
\end{equation*}
\end{example}

@@

\subsection{Extremely large MIS enumeration space}
In many times, the number of configurations is too large to fit into any type of storage.
Then instead of using the set algebra, we can use a binary sum-product tree as a compact representation of the results.
We use a quadruple $(type, data, left, right)$ to represent a node in the tree, where $type$ is one of
$\{\texttt{LEAF}, \texttt{ZERO}, \texttt{SUM}, \texttt{PROD}\}$, $data$ is a bit string as the content is a \texttt{LEAF} node,
$left$ and $right$ are left and right operands of \texttt{SUM} and \texttt{PROD} nodes.

>>>>

%%%%%% Comment related
\newcommand{\red}[1]{[{\bf  \color{red}{ST: #1}}]}
\newcommand{\xred}[1]{[{\bf  \color{red}{\sout{ST: #1}}}]}
\newcommand{\green}[1]{[{\bf  \color{green}{XG: #1}}]}
\newcommand{\xgreen}[1]{[{\bf  \color{green}{\sout{XG: #1}}}]}
\newcommand{\blue}[1]{[{\bf  \color{blue}{JG: #1}}]}
\newcommand{\xblue}[1]{[{\bf  \color{blue}{\sout{JG: #1}}}]}
\newcommand{\cyan}[1]{[{\bf  \color{cyan}{ML: #1}}]}
\newcommand{\xcyan}[1]{[{\bf  \color{cyan}{\sout{ML: #1}}}]}
\newcommand{\material}[1]{\iffalse[{\bf  \color{cyan}{Material: #1}}]\fi}

@@

\begin{abstract}
We introduce a method that uses generic programming tensor network to compute various properties of independent sets,
which include, for example, the size of the maximum independent sets and the number and enumeration of independent sets of a given size.
By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ is the treewidth of the problem graph.
Our framework can be easily extended to compute properties of other problems such as the cut size, coloring, and maximal cliques, among others.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the overlap gap property on three regular graphs.
\end{abstract}

@@

\cyan{a comment} and \xcyan{a resolved comment}

\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting and hard computational problems concerning various properties of independent sets.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$. 
One of the paradigmatic independent set \textit{properties} % that many computational complexity theorists are most interested in 
is the maximum independent set (MIS) and the size of the MIS, $\alpha(G) \equiv \max_{I}|I|$. 
There is widespread interest in this MIS problem not only because it is a well-known NP-complete problem~\cite{Hastad1996} that can be mapped from many other important combinatorial optimization problems in polynomial time such as the 3-satisfiability problem, the maximum clique problem, and the minimum vertex cover problem~\cite{Moore2011},
but also because it has a wide range of applications in scheduling, logistics, wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}.
On the other hand, some other properties, such as the number and enumeration of independent sets or maximal independent sets of a given size, are equally interesting, if not more important, in many practical applications. These other properties can be found in physics applications such as in the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022};
they can, for example, be used to understand phase transitions~\cite{Dyre2016, Fernandes2007}, to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the presence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} related to the clustering phenomenon of large independent sets.
Meanwhile, finding the number of independent sets of a given size is equivalent to computing the coefficients of the independence polynomial~\cite{Harvey2018,Ferrin2014}.
which is a useful graph characteristic related to, for example, the partition functions~\cite{Lee1952,Yang1952} and Euler characteristics of the independence complex~\cite{Bousquet2008, Levit2009}. However, there is a lack of general and versatile tool to compute these different properties of independent sets.
In this paper, we introduce a generic programming tensor-network framework to compute a number of different independent set properties.
We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
It is different from the standard tensor network methods in that the tensor elements are not restricted to standard number types such as floating point numbers and integers.
Traditional algorithms for finding the MIS size include the branching algorithms~\cite{Tarjan1977, Robson1986} and dynamic programming based algorithms~\cite{Courcelle1990, Fomin2013}.
The complexity of our tensor-network based algorithms is similar to that of dynamic programming algorithms~\cite{Courcelle1990, Fomin2013}, which have a complexity scaling exponentially with the treewidth, ${\rm tw}(G)$, of the graph;
for sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.
We benchmark our algorithms in \Sec{sec:benchmark} by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
The high performance benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021},
where efficient heuristic algorithms have been developed to optimize the contraction order for tensor networks up to thousands of vertices.
For cases where the tensors are too large to fit into a GPU, slicing techniques can be used to further reduce the space consumption.
Lastly, we provide a few examples and demonstrate the versatility of our tool by computing the entropy constant for some hardcore lattice gases on 2D square lattices and analyzing the presence of overlap gap property on three regular graphs.
Our method can also be used to find ``maximal'' independent set properties; 
a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum. 
We show how to compute properties related to maximal independent sets and properties for other combinatorial optimization problems in \App{app:otherproblems}.

\section{Tensor networks}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts;
it can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in the subscript to represent a degree of freedom.
One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice.
Hence a tensor network can be represented as a simple graph,
where a tensor is mapped to a vertex and a label shared by two tensors is mapped to an edge.
In this work, we do not impose such a restriction, so an index can appear an arbitrary number of times. 
The graphical representation of a generalized tensor network is a hypergraph, in which an edge (a label in subscript) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{app:tensorbad}.

@@

\section{Generic programming}\label{sec:generic}
In previous works relating tensor networks and combinatoric problems~\cite{Kourtis2019, Biamonte2017},
the elements in the tensor networks are limited to standard number types such as floating point numbers and integers.
Owning to the development of modern compiling technology, we no longer need to limit our imagination to standard number types.
One of the key concepts that push the technology forward is called generic programming:

@@

\section{Independence polynomial} \label{sec:indpoly}
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in graph $G=(V,E)$. The total number of independent sets is thus equal to $I(G, 1)$.
To compute this polynomial, we map this expression to a tensor network contraction.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i)$ as
\begin{equation}
    W(x_i) = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i)_1=x_i$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.
In the special case of $x_i = x$, the contraction result directly corresponds to the independence polynomial.
The connection can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise. The summation counts the number of independent sets of size $k$.

@@

    \centerline{\begin{tikzpicture}[
    dot/.style = {circle, fill, minimum size=#1,
                inner sep=0pt, outer sep=0pt},
    dot/.default = 6pt  % size of the circle diameter 
                    ]  
        \def\dx{0};
        \def\r{0.25cm}
        \filldraw[fill=black] (\dx,0) circle [radius=\r];
        \filldraw[fill=black] (\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,0) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (2.5+\dx,2.5) circle [radius=\r];
        \draw [black,thick] (\dx,0) -- (\dx,1.5);
        \draw [black,thick] (\dx,0) -- (1.5+\dx,0);
        \draw [black,thick] (\dx,1.5) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (\dx,1.5);
        \draw [black,thick] (2.5+\dx,2.5) -- (1.5+\dx,1.5);
        \node[color=white] at (\dx,0) {a};
        \node[color=white] at (\dx,1.5) {b};
        \node[color=white] at (1.5+\dx,0) {c};
        \node[color=white] at (1.5+\dx,1.5) {d};
        \node[color=white] at (2.5+\dx,2.5) {e};
        \def\dx{5};
        \def\r{0.25cm}
        \foreach \x/\y/\e in {0.75/0/ac, 0/0.75/ab, 1.5/0.75/cd, 0.75/1.5/bd, 0.75/0.75/bc, 2/2/de}
            \node[color=white,fill=black,dot=2*\r] at (\x+\dx,\y) (\e) {\scriptsize B};
        \foreach \x/\y/\v in {0/0/a, 0/1.5/b, 1.5/0/c, 1.5/1.5/d, 2.5/2.5/e}
            \node[color=black] at (\x+\dx,\y) (\v) {$s_\v$};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=white,fill=black,dot=\r] at (\x+\dx,\y) (\v\v) {};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=black] at (\x+\dx+0.5,\y) {\scriptsize $W(x_\v)$};
        \draw [cyan,thick] (a) -- (aa);
        \draw [cyan,thick] (a) -- (ab);
        \draw [cyan,thick] (a) -- (ac);
        \draw [blue,thick] (b) -- (bb);
        \draw [blue,thick] (b) -- (ab);
        \draw [blue,thick] (b) -- (bc);
        \draw [blue,thick] (b) -- (bd);
        \draw [red,thick] (c) -- (cc);
        \draw [red,thick] (c) -- (ac);
        \draw [red,thick] (c) -- (bc);
        \draw [red,thick] (c) -- (cd);
        \draw [green,thick] (d) -- (dd);
        \draw [green,thick] (d) -- (bd);
        \draw [green,thick] (d) -- (de);
        \draw [green,thick] (d) -- (cd);
        \draw [orange,thick] (e) -- (ee);
        \draw [orange,thick] (e) -- (de);
    \end{tikzpicture}}

    The contraction of this tensor network can be done in a pairwise order utilizing the associativity, additive commutativity, and multiplicative commutativity of tensor elements:
    \begin{align*}
        &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a)_{s_a} W(x_b)_{s_b} W(x_c)_{s_c} W(x_d)_{s_d} W(x_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_c s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}\\
        =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e)_{s_e}\right) W(x_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c)_{s_c}\right)\right)\right.\\
        &\phantom{XXX}\left.\left(B_{s_bs_c}\left(\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a)_{s_a}\right)\right)\right)\right)\\
        =&1 + x_a + x_b + x_c + x_d + x_e + x_ax_d + x_ax_e + x_cx_e + x_bx_e\\
        =&1+5x+4x^2 \qquad \quad (x_{i} = x).
    \end{align*}
\end{example}

Before contracting the tensor network and evaluating the independence polynomial numerically, let us first elevate the tensor elements $0$s and $1$s in tensors $W(x)$ and $B$ from integers and floating point numbers to the additive identity,

@@

However, using the polynomial type suffers a space overhead proportional to $\alpha(G)$ because each polynomial requires a vector of such size to store the coefficients. 

@@

\section{Enumeration of independent sets}
\subsection{Set algebra for configuration enumeration}
The enumeration problems of independent sets are also interesting and has been studied extensively in the literature~\cite{Bron1973, Eppstein2010, Johnson1988}, including,
for example, the enumeration of all independent sets, the enumeration of all maximal independent sets, or the enumeration of all MISs.
To enumerate all independent sets, we designed an algebra defined on sets of bitstrings:
\begin{equation}
\eqname{SN}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau \, | \, \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}.
\end{split}\label{eq:set}
\end{equation}
where $s$ and $t$ are each a set of $|V|$-bit strings and $\lor^\circ$ is the bitwise OR operation over two bit strings.
\begin{example}\label{eg:setalgebra}
    For elements being bit strings of length $5$, we have the following set algebra
\begin{equation*}
\begin{split}
    &\{00001\} \oplus \{01110, 01000\} = \{01110, 01000\} \oplus \{00001\} = \{00001,01110, 01000\}\\
    &\{00001\} \oplus \{\} = \{00001\}\\
&\\
    &\{00001\} \odot \{01110, 01000\} = \{01110, 01000\} \odot \{00001\} = \{01111, 01001\}\\
    &\{00001\} \odot \{\} = \{\}\\
    &\{00001\} \odot \{00000\} = \{00001\}.
\end{split}
\end{equation*}
\end{example}

@@

\section{Benchmarks and case studies}\label{sec:benchmark}
\subsection{Performance benchmarks}
We run a single thread benchmark on CPU Intel(R) Xeon(R) CPU E5-2686 v4 @ 2.30GHz,
and its CUDA version on a GPU Tesla V100-SXM2 16G.
The results are summarized in Figure~\ref{fig:benchmark}.
The graphs that we use in benchmarks are random three regular graphs,
 a typical type of sparse graphs that has a small treewidth that asymptotically smaller than $|V|/6$~\cite{Fomin2006}.
